Application of Generalized Additive Models for long-term and seasonal trend analysis of water quality: South San Francisco Bay case study

### 2020-05-28 manu_draft.Rmd
added fig 1, table 1

---
title: "Application of Generalized Additive Models for long-term and seasonal trend analysis of water quality: South San Francisco Bay case study"
output: 
  bookdown::word_document2:
    reference_docx: my_styles.docx
bibliography: refs.bib
author: "Marcus W. Beck (mbeck@tbep.org), Perry de Valpine (pdevalpine@berkeley.edu), Rebecca Murphy (rmurphy@chesapeakebay.net), Ian Wren (ianw@sfei.org), Ariella Chelsky (ariellac@sfei.org), Melissa Foley (melissaf@sfei.org), David B. Senn (davids@sfei.org)"
urlcolor: blue
csl: ecology.csl
link-citations: true
---

@@

For each GAM, chlorophyll is modelled using a link function $g()$ that relates the response variable to the independent variables $year$ and day of year ($doy$).  The $year$ variable is a continuous measure for date in decimal time, where date is expressed as a continuous numeric value with year on the left of the decimal and date within the year to the right of the decimal (e.g., July 1st 2019 would be 2019.5).  The $doy$ variable is an integer value for the Julian day in each year (from 1 to 366) that represents the seasonal component of the time series. The estimate $s_0$ is an intercept value and $\beta_1$ is a slope estimate for a linear year effect included in all models. The $s_1()$ and $s_2()$ functions are different smoothers for $year$ and $doy$ where the former uses a thin-plate spline and the latter uses a cyclic cubic spline to capture the periodic seasonal component.  The $ti_1()$ function is a tensor product interaction between $year$ and $doy$ that allows the seasonal component to change between years.  Within the $ti_1()$ function are the two separate smoothers for year as a thin-plate spline and $doy$ as a cyclic cubic spline. As with conventional linear models, the remaining variance in the response not captured by the functional form of the model is expressed as an error term $\epsilon$ that is normally-distributed having mean equal to zero and variance as $\sigma^2$.  The residual variance is used to back-transform model estimates for trend testing, described below.

@@

A challenge for optimizing GAMs to time series data with significant intra- and interannual variation is choosing an appropriate functional form (i.e., gam0, gam1, gam2, or gam6) and determining how much variation could be explained by the smoothers in each function.  The four models above provide tradeoffs in the functional forms that balance descriptive and computational efficiency.  However, the individual smoothing functions (i.e, $s_1()$, $s_2()$, $s_2^*()$ and $ti_1()$) require an explicit declaration of how much variation could be explained for the separate $year$ and $doy$ components.  This requires user input on the potential upper limit for the *k*-values that could be used during model fitting.  Initial testing by Murphy et al. (2019) showed that the default *k* values in the mgcv package were insufficient for describing the interannual variation in the response variable and specified the upper limit for the number of knots for the $s_2(year)$ smoother as the maximum between 10 or  $2 / 3 \times$ the number of years in the time series.   We follow the same approach herein for gam1 and gam2.  Further, the default *k*-values for $s_2()$ and $ti_1()$ were considered appropriate and not tested further. For the application to SFE chlorophyll data, gam6 was also added as an extension of gam1, where the total number of knots for the $year$ smoother, $s_2^*()$, was increased to 12 times the total number of years in the time series.  This approach potentially allows the smoother for $year$ to be completely unconstrained in how much variance is fit within and between years.

@@

@@

>>>>

---
title: "Application of Generalized Additive Models for long-term and seasonal trend analysis of water quality: South San Francisco Bay case study"
output: 
  bookdown::word_document2:
    reference_docx: my_styles.docx
always_allow_html: true
bibliography: refs.bib
author: "Marcus W. Beck (mbeck@tbep.org), Perry de Valpine (pdevalpine@berkeley.edu), Rebecca Murphy (rmurphy@chesapeakebay.net), Ian Wren (ianw@sfei.org), Ariella Chelsky (ariellac@sfei.org), Melissa Foley (melissaf@sfei.org), David B. Senn (davids@sfei.org)"
urlcolor: blue
csl: ecology.csl
link-citations: true
---

@@

For each GAM, chlorophyll is modelled using a link function $g()$ that relates the response variable to the independent variables $year$ and day of year ($doy$) (Table \@ref(tab:modsumtab), Figure \@ref(fig:modsumfig)).  The $year$ variable is a continuous measure for date in decimal time, where date is expressed as a continuous numeric value with year on the left of the decimal and date within the year to the right of the decimal (e.g., July 1st 2019 would be 2019.5).  The $doy$ variable is an integer value for the Julian day in each year (from 1 to 366) that represents the seasonal component of the time series. The estimate $s_0$ is an intercept value and $\beta_1$ is a slope estimate for a linear year effect included in all models. The $s_1()$ and $s_2()$ functions are different smoothers for $year$ and $doy$ where the former uses a thin-plate spline and the latter uses a cyclic cubic spline to capture the periodic seasonal component.  The $ti_1()$ function is a tensor product interaction between $year$ and $doy$ that allows the seasonal component to change between years.  Within the $ti_1()$ function are the two separate smoothers for year as a thin-plate spline and $doy$ as a cyclic cubic spline. As with conventional linear models, the remaining variance in the response not captured by the functional form of the model is expressed as an error term $\epsilon$ that is normally-distributed having mean equal to zero and variance as $\sigma^2$.  The residual variance is used to back-transform model estimates for trend testing, described below.

@@

A challenge for optimizing GAMs to time series data with significant intra- and interannual variation is choosing an appropriate functional form (i.e., gam0, gam1, gam2, or gam6) and determining how much variation could be explained by the smoothers in each function.  The four models above provide tradeoffs in the functional forms that balance descriptive and computational efficiency (Table \@ref(tab:modsumtab), Figure \@ref(fig:modsumfig)).  However, the individual smoothing functions (i.e, $s_1()$, $s_2()$, $s_2^*()$ and $ti_1()$) require an explicit declaration of how much variation could be explained for the separate $year$ and $doy$ components.  This requires user input on the potential upper limit for the *k*-values that could be used during model fitting.  Initial testing by Murphy et al. (2019) showed that the default *k* values in the mgcv package were insufficient for describing the interannual variation in the response variable and specified the upper limit for the number of knots for the $s_2(year)$ smoother as the maximum between 10 or  $2 / 3 \times$ the number of years in the time series.   We follow the same approach herein for gam1 and gam2.  Further, the default *k*-values for $s_2()$ and $ti_1()$ were considered appropriate and not tested further. For the application to SFE chlorophyll data, gam6 was also added as an extension of gam1, where the total number of knots for the $year$ smoother, $s_2^*()$, was increased to 12 times the total number of years in the time series.  This approach potentially allows the smoother for $year$ to be completely unconstrained in how much variance is fit within and between years.

@@

```{r modsumfig, fig.height = 4.5, fig.width = 6, fig.cap = "General output for each of the four GAM structures in equations \\@ref(eq:gamzero) to \\@ref(eq:gamsix).  Day of year (Julian) is on the x-axis and each curve represents a different fitted year for chlorophyll concentration."}
# get predictions for all four gams
tomod <- rawdat %>%
  filter(station %in% 22) %>%
  filter(param %in% 'chl')

# use previously fitted list of models
trans <- 'boxcox'
mods <- list(
  gam0 = anlz_gam(tomod, mod = 'gam0', trans = trans),
  gam1 = anlz_gam(tomod, mod = 'gam1', trans = trans),
  gam2 = anlz_gam(tomod, mod = 'gam2', trans = trans),
  gam6 = anlz_gam(tomod, mod = 'gam6', trans = trans)
)

p1 <- show_prddoy(mods = mods, ylab = 'Chlorophyll-a (ug/L)', nfac = 2, size = 0.8, alpha = 0.8) + 
  theme(legend.position = 'right') + 
  guides(colour = ggplot2::guide_colourbar(barheight = 15, barwidth = 1)) +
  scale_colour_gradient(low = "white", high = "black") + 
  theme_minimal() + 
  theme(
    legend.title = element_blank()
  )

p1
```

@@

```{r modsumtab}
tab <- tibble(
  GAM = c('`gam0`', '`gam1`', '`gam2`', '`gam6`'),
  Summary = c(
    'Linear trend with seasonality', 
    'Nonlinear trends seasonality (constrained knots)', 
    'Nonlinear trends wiht seasonality, plus interaction',
    'Nonelinear trends with seasonality (unconstrained knots)'
  ),
  Details = c(
    'A linear trend in time over the course of the time series (cyear); and Within-year seasonal fluctuations that follow the same pattern every year - e.g. large spring bloom and small fall bloom (`s(doy, bs = "cc")`)',
    'Treats cyear and doy same as gam0; and A smooth non-linear trend through time, such that the seasonal peaks are similar across years, but interannual change is evident', 
    'Treats cyear and doy same as gam1; and Within-year seasonal fluctuations that vary across years but in a constrained pattern compared to gam6 (`ti(year, doy, bs = c("tp", "cc"))`)', 
    'Variant of gam1 with much higher values of k, allowing the splines to follow greater fluctuations both within and across years.')
  )

# table stuff
cap.val <- 'Summary and details for each of the GAM structures in equations \\@ref(eq:gamzero) to \\@ref(eq:gamsix).'

# table
knitr::kable(tab, booktabs = T, caption = cap.val)
```
